---
title: Feature Engineering
author: Drew Galloro
format:
  html:
   theme: cosmo
   toc: true
   code-copy: true
   code-block-background: true
   code-overflow: wrap
embed-resources: true
---

# Testing Feature Engineering

## Adding Features from Previous EDA's

Let's start by adding in the weekday/weekend feature, and the pre/post-covid feature from eda1.

```{r}
#| echo: false
#| warning: false
library(tidyverse)
od_data <- read_csv("../data/od-data.csv")
```

```{r}
od_data <- od_data |> mutate(post_covid = if_else(date > ymd("2020-03-19"), 1, 0),
                                day = case_when(wday(date) %in% seq(2, 6) ~ "weekday",
                                                wday(date) == 1 ~ "sunday",
                                                wday(date) == 7 ~ "saturday")
                        )
```

Let's run some code to see what the results are

```{r}
od_data |> 
    group_by(post_covid) |>
    summarize(count = n(), median = median(ridership), mean = mean(ridership))
```

Let's look at the distribution of total ridership faceted by our covid indicator

```{r}
od_data |> 
    ggplot(aes(x = ridership)) + 
        geom_histogram(bins = 30, color = "steelblue") + 
        labs(x = "Ridership",
            y = "Counts") + 
        facet_wrap(~ post_covid)
```

I realized that I can use an ipynb file with an R kernel. I've switched to ipynb files...
