---
title: STAT 230A Project Proposal Draft
subtitle: Predicting BART Ridership Using Hourly OD Data
author:
    - Andrea (Drew) Galloro
    - Benjamin Khothsombath
format:
  pdf:
    documentclass: article
    margin-left: 1in
    margin-right: 1in
    margin-top: 1in
    margin-bottom: 1in
    #toc: true
    fontsize: 12pt
    include-in-header: 
        text : |
            \usepackage{mathtools}
            \usepackage{xcolor}
            \usepackage{listings}
            \usepackage{subcaption}
            \lstset{breaklines=true,breakatwhitespace=true}
            \newcommand{\iid}{\overset{\text{iid}}{\sim}}
            \newcommand{\E}[1]{\mathbb{E}\left[ #1 \right]}
            \newcommand{\Var}[1]{\text{Var}\left( #1 \right)}
            \newcommand{\Prob}[1]{\mathbb{P}\left( #1 \right)}
            \newcommand{\ind}[1]{\mathbb{1}\left\{ #1 \right\}}
            \newcommand{\Cov}[1]{\text{Cov}\left( #1 \right)}
            \newcommand{\sand}{\quad \text{and} \quad}
---

```{r read-in-data}
#| include: false
library(tidyverse)
od_data <- read_csv("../../data/od-data.csv")
set.seed(230)
```

# Introduction

Bay Area Rapid Transit (BART) services connects people from all parts of the San Francisco Bay Area. Thousands of people ride BART every day of the year. However, changes in demographics and a large, sudden shock to public transit ridership have placed BART on the backfoot. Even while they continue to expand service to the San Jose area and furthur inland into Contra Costa county. Building a model to predict ridership from existing data would be an aid to the agency, and could be used to help set schedules, justify new expansion or housing near stations, or justify reduced service in wake of an upcoming budget deficit. Our goal is to take publicly available BART hourly ridership data, organize by date and origin and destination (OD) pairs, and use it to construct a linear model that effectively captures trends in ridership over time.

The dataset we chose to work with is large, but sparce in features. We have `r dim(od_data)[1]` observations in the dataset, but only `r dim(od_data)[2]` features. As a result, our plan is to use these features to engineer new ones that can extract patterns in the data, and use linear modelling principles to evaluate the significance of our engineered features in predicting ridership. How does ridership vary by different features, and to what degree has BART been able to bounce back after the COVID-19 pandemic?

# The Data

Our data is sourced from BART's ridership reports: [Ridership Reports](https://www.bart.gov/about/reports/ridership). Specifically, the hourly trip datasets, which run from 2018 to 2025 (The 2026 dataset has not been updated yet). An observation in our data looks like the following:

```{r make-observation-table}
#| echo: false
knitr::kable(slice_sample(od_data, n = 1))
```

We have the date, hour (using 24-hour clock), the origin station, destination station (both represented as four letter codes), and the ridership. This specific observation is from May 24th, 2018, and tells us that 1 person rode from the San Bruno station to the 19th St. Oakland station during the 3:00 - 4:00 PM hour. Our assumption is that this means this person exited at the 19th St Mission station during this time period, as exiting the BART system would create the OD pair, but that is not explicitly stated anywhere in the official data description. 

### Ridership Distributions

There are $50*49 + 50 = 2500$ different possible OD pairs in the data. BART has 50 stations, order of OD data matters, and there are instances of people getting on and off at the same station. Below are histograms of ridership numbers for a random sample of 6 OD pairs:

```{r add-pairs-for-faceting}
#| echo: false
#| warning: false
od_data <- od_data |> mutate(pair = paste(origin, destination, sep = " -> "))
stations <- od_data |> 
                distinct(pair) |>
                slice_sample(n = 6)
```

```{r plot-ridership-histogram}
#| echo: false
#| warning: false
od_data |> 
    filter(pair %in% stations$pair) |>
                ggplot() + 
                        geom_histogram(aes(x = ridership), bins = 30) + 
                        facet_wrap(~ pair) + 
                        labs(title = "Ridership for Random Sample of OD Pairs",
                            x = "Ridership",
                            y = "Counts")
```

We can see here that ridership is a right-skewed output variable, and there are observations with relatively low numbers. The data does not include datetime-OD data for OD pairs that did not not have any riders, so we are safe to do a log-transform of the ridership variable.

### Ridership Patterns

We can also look at trends in ridership over the course of the day. Below are plots of ridership per hour for the same six OD pairs, limiting the data to just weekdays.

```{r plot-hourly-ridership}
#| echo: false
#| warning: false
od_data |> 
    filter(pair %in% stations$pair & 
            wday(date) %in% seq(2, 6)) |>
    group_by(pair, hour) |>
    summarize(riders = sum(ridership)) |>
                ggplot(aes(x = hour, y = riders)) + 
                        geom_line() + 
                        facet_wrap(~ pair) + 
                        labs(title = "Ridership Patterns by Hour for Random Sample of OD Pairs",
                            x = "Hour of Day",
                            y = "Ridership")
```

We can see again that there are some OD pairs that go relatively unused, such as South San Francisco (SSAN) to Castro Valley (CAST). On the other hand, the ones that do have relatively high ridership, such as Embarcadero (EMBR) to Civic Center (CIVC), exhibit the same pattern on Weekdays. Unsurprisingly, riderhip exhibits spikes at the beginning and end of the workday. It's important to note again that order of OD pairs matters in the dataset. These plots do not display riders who are going the other direction.

### Plot Exit Rates at Stations

Let's plot the exit rates for a October 15th, 2025 on a map of the SF Bay Area: 

__This is a fun idea, but i would need to implement some geospatial data as well and I don't know if I have time for that right now.__ 

```{r make-map}
# eval: false
# echo: false

# would need to add geospatial data to do this
od_data |> 
    filter(date == ymd("2025-10-15")) |>
    group_by(destination) |>
    summarize(exit_rate = n())
```

# Modelling Plan

